Temperature and wave exposure as drivers of morphology in two kelp species

### 2019-07-31 morphology_paper.Rmd
updated project options. added in results section for boxplots and rda outputs completed

---
title: Temperature and wave exposure as drivers of morphology in two kelp species
  along the coast of South Africa
author: "R. Coppin, C. Rautenbach, T. Ponton and A.J Smit"
geometry: margin=1in
output:
  pdf_document:
    fig_caption: yes
    fig_crop: no
    fig_height: 7
    fig_width: 7
  html_notebook:
    fig_caption: yes
    fig_height: 7
    fig_width: 7
  html_document:
    df_print: paged
  word_document:
    reference_docx: morphology_paper_files/frontiers_template.docx
papersize: A4
tables: yes
fontsize: 10pt
---

@@

Sites were chosen to represent an array of environmental gradients, as indicated in Fig. 1. St. Helena Bay and Betty's Bay constituted the north western and south eastern boundary sites, respectively. These sites are roughly 300 km apart, and lie within separate marine provinces[3] and cover a cline in wave exposure and temperature. Study sites  span across the majority of the south-west coast, in varying thermal and wave energy regimes. The region is dominated by kelp communities that persist in contrasting abiotic environments. The west coast region has been termed as a cool temperate region, which is defined as a region where mean monthly temperatures are always above 10°C and always below 15°C (Smit et al. 2013). East of Cape Point marks the beginning of an overlap or transition area, which is also referred to as the Benguela-Agulhas Transition Zone (Smit et al. 2013). The Agulhas Marine Province is characterised by a wide temperature range of up to 7°C difference between mean monthly temperatures between summer and winter (Smit et al. 2013). Due to the Cape Peninsula temperate latitude, winter months bring an increased frequency of frontal depressions that originate from the Southern Ocean (Reason et al. 2006). These low pressures are joined by large swells with increased wave energy. The nearshore environment, with the accompanied biota, therefore experiences high wave energy events, with increased frequency in winter (Veitch et al. 2018). The large peninsula acts as an obstruction for large south westerly swells, providing decreased wave energy along the west side of False Bay (Shipley 1964). Conversely, the west coast of Cape Point is battered by these large swells. Multiple sites, therefore, exist where kelps grow in diverse temperature and wave energy climates, in close proximity.

@@

## Abiotic environment
### Temperature

In order to compare abiotic variables for sites around the coast, large historical databases for temperature, wave energy and wind were accessed. Shallow water temperatures were sourced from The South African Coastal Temperature Network (SACTN) website (https://github.com/ajsmit/SACTN). In terms of nearshore temperature, in situ data are preferred over satellite SST, which have shown to exhibit large biases (Smit et al. 2013). Linear interpolated SST were calculated for sites where *in situ* recorders were absent. These data were used to group sites into 'Cool temperate' and 'Warm temperate' categorisations of thermal regime.

### wave environment

Wave model variable data formed part of the South African Coastal Vulnerability Assessment, presented to the Department of Environmental Affairs (DEA) and produced by the Council for Scientific and Industrial Research (CSIR) (Rautenbach 2015). The South African wave climate was modeled via 20 spectral numerical wave models that simulated the offshore wave climate to the nearshore. The boundary conditions of these models were obtained by using the NOAA Wave Watch III (WWIII) model output, distributed via the National Centers for Environmental Prediction (NCEP) product (Office of the Director 2000; Environmental Modeling Center / Marine Modeling Branch 2005). The particular hindcast product utilised during the DEA-CSIR study spans 1994-2013 at a 3-hour resolution. These data were then used to model swell propagation into the coastal models while wind waves (seas) were generated via stationary computations in the Simulating Waves in the Nearshore SWAN model (Booij et al. 1997). The assumption of stationary computations are acceptable as the model domains were small enough so the temporal variation of the model boundary were slower than the time it takes for that boundary condition to propagate to the coast.  SWAN allows one to extract wave variables from specific gridded locations in the nearshore. For False Bay, a resolution of 200 meters was modelled and output produced at both the 7 meter and 15 meter isobaths. A 200 meter resolution was used as the False Bay computational grid was nested within a larger grid (1 kilometer resolution). This allowed for a computational effective wave resolution of increasing resolution, from the NCEP, low resolution output to nested, high resolution coastal output. For Table Bay and east of Cape Hangklip the resolution was 500 meters and also had output at the 7 meters and 15 meter isobaths. These contour outputs were chosen in the original study by the CSIR as most engineering run-up calculations require wave parameter information at these contour depths and were the main focus of the original study. For this study the 7 meter contours were used. These data were used to group sites according to

@@

# Results
## Abiotic environment

@@

@@

```{r wave variable processing, warning=FALSE, include=FALSE, echo = FALSE, message=FALSE}
### Plots for wave variables
#Annual wave plots

positions_wave <- c("Betty's Bay", "Roman Rock","Kalk Bay", "A-Frame", "Miller's Point", "Baboon Rock",  "Batsata Rock","Bordjies Deep","Buffelsbaai","Buffelsbaai South","Olifantsbos","Scarborough", "Soetwater", "Kommetjie","Hout Bay","Oudekraal","Sea Point","St. Helena Bay")

@@

**Insert wave exposure maps here**

@@

@@

```{r laminaria temp plots, echo=FALSE, warning=FALSE, fig.cap='Boxplots of *L. pallida* morphological characteristics group by temperature regime.', message=FALSE, warning=FALSE, include=FALSE}
ggarrange(LT1, LT2,LT3,LT4,LT5,LT6,LT7,LT8,LT9, labels = c("AUTO"), common.legend = FALSE, legend = "bottom")
```

```{r lamanaria wave plots, echo=FALSE,warning=FALSE, fig.cap='Boxplots of *L. pallida* morphological characteristics group by wave exposure category.'}
ggarrange(LW1, LW2,LW3,LW4,LW5,LW6,LW7,LW8,LW9, labels = c("AUTO"), common.legend = FALSE, legend = "bottom")
```

@@

```{r deep ecklonia temp plot, message=FALSE, warning=FALSE, fig.cap ='Boxplot of morphological characteristics for deep population *E. maxima* grouped by temperature.'}
ggarrange(ET1,ET2,ET3,ET4,ET5,ET6,ET7,ET8,ET9,ET10, labels = "AUTO", common.legend = FALSE, legend = "bottom")
```

```{r deep ecklonia wave plots, warning=FALSE,fig.cap ='Boxplot of morphological characteristics for deep population *E. maxima* grouped by wave exposure.'}
ggarrange(EW1,EW2,EW3,EW4,EW5,EW6,EW7,EW8,EW9,EW10, labels = "AUTO", common.legend = FALSE, legend = "bottom")
```

>>>>

---
title: Temperature and wave exposure as drivers of morphology in two kelp species
  along the coast of South Africa
author: "R. Coppin, C. Rautenbach, T. Ponton and A.J Smit"
geometry: margin=1in
output:
  html_notebook:
    fig_caption: yes
    fig_height: 7
    fig_width: 7
  pdf_document:
    fig_caption: yes
    fig_crop: no
    fig_height: 7
    fig_width: 7
  html_document:
    df_print: paged
  word_document:
    reference_docx: morphology_paper_files/frontiers_template.docx
papersize: A4
tables: yes
fontsize: 10pt
---

@@

Sites were chosen to represent an array of environmental gradients, as indicated in Fig. 1. St. Helena Bay and Betty's Bay constituted the north western and south eastern boundary sites, respectively. These sites are roughly 300 km apart, and lie within separate marine provinces[3] and cover a cline in wave exposure and temperature. Study sites span across the majority of the south-west coast, in varying thermal and wave energy regimes. The region is dominated by kelp communities that persist in contrasting abiotic environments. The west coast region has been termed as a cool temperate region, which is defined as a region where mean monthly temperatures are always above 10°C and always below 15°C (Smit et al. 2013). East of Cape Point marks the beginning of an overlap or transition area, which is also referred to as the Benguela-Agulhas Transition Zone (Smit et al. 2013). The Agulhas Marine Province is characterised by a wide temperature range of up to 7°C difference between mean monthly temperatures between summer and winter and is classified as warm temperate region (Smit et al. 2013). Due to the Cape Peninsula temperate latitude, winter months bring an increased frequency of frontal depressions that originate from the Southern Ocean (Reason et al. 2006). These low pressures are joined by large swells with increased wave energy. The nearshore environment, with the accompanied biota, therefore experiences high wave energy events, with increased frequency in winter (Veitch et al. 2018). The large peninsula acts as an obstruction for large south westerly swells, providing decreased wave energy along the west side of False Bay (Shipley 1964). Conversely, the west coast of Cape Point is battered by these large swells. Multiple sites, therefore, exist where kelps grow in diverse temperature and wave energy climates, in close proximity.

@@

## Abiotic environment

In order to compare abiotic variables for sites around the coast, large historical databases for temperature, wave energy and wind were accessed.

### Temperature

Shallow water temperatures were sourced from The South African Coastal Temperature Network (SACTN) website (https://github.com/ajsmit/SACTN). In terms of nearshore temperature, *in situ* data are preferred over satellite SST, which have shown to exhibit large biases (Smit et al. 2013). Linear interpolated SST were calculated for sites where *in situ* recorders were absent. These data were used to group sites into 'Cool temperate' and 'Warm temperate' categorisations to investigate the possible affect of thermal regime on kelp morphological characteristics.

### Wave environment

Wave model variable data formed part of the South African Coastal Vulnerability Assessment, presented to the Department of Environmental Affairs (DEA) and produced by the Council for Scientific and Industrial Research (CSIR) (Rautenbach 2015). The South African wave climate was modeled via 20 spectral numerical wave models that simulated the offshore wave climate to the nearshore. The boundary conditions of these models were obtained by using the NOAA Wave Watch III (WWIII) model output, distributed via the National Centers for Environmental Prediction (NCEP) product (Office of the Director 2000; Environmental Modeling Center / Marine Modeling Branch 2005). The particular hindcast product utilised during the DEA-CSIR study spans 1994-2013 at a 3-hour resolution. These data were then used to model swell propagation into the coastal models while wind waves (seas) were generated via stationary computations in the Simulating Waves in the Nearshore SWAN model (Booij et al. 1997). The assumption of stationary computations are acceptable as the model domains were small enough so the temporal variation of the model boundary were slower than the time it takes for that boundary condition to propagate to the coast.  SWAN allows one to extract wave variables from specific gridded locations in the nearshore. For False Bay, a resolution of 200 meters was modelled and output produced at both the 7 meter and 15 meter isobaths. A 200 meter resolution was used as the False Bay computational grid was nested within a larger grid (1 kilometer resolution). This allowed for a computational effective wave resolution of increasing resolution, from the NCEP, low resolution output to nested, high resolution coastal output. For Table Bay and east of Cape Hangklip the resolution was 500 meters and also had output at the 7 meters and 15 meter isobaths. These contour outputs were chosen in the original study by the CSIR as most engineering run-up calculations require wave parameter information at these contour depths and were the main focus of the original study. For this study the 7 meter contours were used. These data were then used to calculate over all wave power (kW/m) as this was considered the best measure of overall wave exposure. Annual wave power was plotted and categorised into different wave exposure categories which ranged from fully sheltered to extremely exposed.

@@

# Results

## Abiotic environment

@@

Our data show the western side of the peninsula experiences higher significant wave heights and variation in wave heights compared to the eastern side of the peninsula on both an annual and seasonal scale. Annual mean significant wave height ranged from 0.5 ± 0.2 m (mean ± SD) at St. Helena Bay to 2.3  ± 0.7 m at Olifantsbos on the western side of the peninsula, and on the eastern side of the peninsula it ranged from 0.9 ± 0.4 m at Buffelsbaai to 1.9 ± 0.6 m at Betty’s Bay. Annual mean wave peak period for sites on the western side of the peninsula ranged from 10.3 ± 3.5 s at St. Helena Bay to 11.0 ± 2.0 s at Olifantsbos, and ranged from 10.6 ± 3.0 s at Buffelsbaai to 10.8 ± 2.2 s at Betty’s Bay. The data shows no trend in mean peak wave period for the coastline. Annual peak wave period was 10.0 ± 3.0 s and 9.2 ± 3.1 s in winter and summer respectively, Miller’s point and had the lowest peak wave period amongst all sites. These data show that the western side of the peninsula has a lower variation (SD) compared to the eastern side of the peninsula, with Miller’s point showing the highest variation across timescales.  Annual maximum peak wave period ranged from 18.9 s at St. Helena Bay to 18.4 s at Olifantsbos for the western side of the peninsula, and ranged from 18.0 s at Buffelsbaai to 18.0 s at Betty’s Bay on the eastern side of the peninsula.

@@

```{r wave variable processing, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
### Plots for wave variables
#Annual wave plots

positions_wave <- c("Betty's Bay", "Roman Rock","Kalk Bay", "A-Frame", "Miller's Point", "Baboon Rock",  "Batsata Rock","Bordjies Deep","Buffelsbaai","Buffelsbaai South","Olifantsbos","Scarborough", "Soetwater", "Kommetjie","Hout Bay","Oudekraal","Sea Point","St Helena Bay")

@@

```{r wave power map, echo=FALSE, fig.cap='Schematic of *E. maxima* on the left and *L. pallida* on the right. (Dyer 2018)'}
library(png)
library(grid)

img_wavepower <- readPNG("figures/CapePen_Total.PNG")
grid.raster(img_wavepower)
```

@@

Significant differences were found between cool temperate and warm temperate regimes for all *L. pallida* morphological differences with lamina length, number of digits and thallus mass the exceptions. Certian morphological characteristics such as stipe mass, stipe length, total length and stipe diameter had significanlty higher means for the cool temperate regimes when compared to the warm temperate regime. Conversely, lamina weight and lamina thickness has significantly lower means for the cool temperate regime compared to the warm temperate regime. When *L. pallida* morphological characteristics were grouped to wave exposure categories kruskal-wallis tests revealed significant differences amoungst categories for all morphological characteristics with lamina length, number of digits and thallus mass the exceptions. The morphological characteristics that were significantly different amoungst sites exibited similar patterns, with increasing mean values from the full sheltered to extremely sheltered categories and then a decrease in mean values for the remainder of the categories. Two morphological characteristics, namely lamina thickness and stipe diameter, both exibit higher variations for the exposed categories.

@@

```{r laminaria temp plots, echo=FALSE, warning=FALSE, fig.cap='Boxplots of *L. pallida* morphological characteristics group by temperature regime.', message=FALSE, warning=FALSE, include=FALSE, fig.height=8, fig.width=7}
ggarrange(LT1, LT2,LT3,LT4,LT5,LT6,LT7,LT8,LT9, labels = c("AUTO"), common.legend = TRUE, legend = "bottom")
```

```{r lamanaria wave plots, echo=FALSE,warning=FALSE, fig.cap='Boxplots of *L. pallida* morphological characteristics group by wave exposure category.', fig.height=8, fig.width=8}
ggarrange(LW1, LW2,LW3,LW4,LW5,LW6,LW7,LW8,LW9, labels = c("AUTO"), common.legend = TRUE, legend = "bottom")
```

All deep *E. maxima* morphological characteristics showed differences between cool temperate and warm temperate regimes except for frond mass, frond length and stipe mass. Mean values for primary length, primary width and stipe circumference were all significantly higher for the cool temperate regime than the warm temperate regime. The remaining morphological characteristics which showed significant differences between regimes exhibited significanlty lower values for the cool temperate regime compared to the warm temperate regime. Significant differences amoungst wave exposure categories were found for all morphological characteristics of deep *E. maxima* populations. Certian morphological characteristics showed a gradual increase in mean value with inncreasing wave exposure. The remaining morphological characteristics exhibited decreased values for only the higher exposure categories and then a sharp increase in value for the highest wave exposure category. In all cases, mean values for the highest wave exposure level (very exposed) was greatest compared to other levels of exposure.

@@

```{r deep ecklonia temp plot, message=FALSE, warning=FALSE, fig.cap ='Boxplot of morphological characteristics for deep population *E. maxima* grouped by temperature.',fig.height=8, fig.width=8}
ggarrange(ET1,ET2,ET3,ET4,ET5,ET6,ET7,ET8,ET9,ET10, labels = "AUTO", common.legend = TRUE, legend = "bottom")
```

```{r deep ecklonia wave plots, warning=FALSE,fig.cap ='Boxplot of morphological characteristics for deep population *E. maxima* grouped by wave exposure.',fig.height=8, fig.width=8}
ggarrange(EW1,EW2,EW3,EW4,EW5,EW6,EW7,EW8,EW9,EW10, labels = "AUTO", common.legend = TRUE, legend = "bottom")
```
